Reasonably Programmable Syntax

### 2017-05-30 popl18/syntax-popl18.tex
tweaks to sec 1.0 and 1.1

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to build in literal forms for a select few standard data structures, e.g. lists. The problem is that this practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.) Instead, they must reason transparently, i.e. about the underlying expansion. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.

This paper formally details a more reasonable alternative: \emph{typed literal macros} (TLMs). TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning 1) that expansions cannot make any assumptions about the application site context; and 2) that expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced segments. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.  This is the first detailed type-theoretic account of a typed, hygienic syntactic macro system for an ML-like language. % This calculus is the first such formalization of a typed macro system.

@@

For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the expansion of literal forms is \emph{context independent}. We will return to the concept of context independence throughout this work.} List pattern literals are analagous. Another example is Ur/Web, which defines a dialect of Ur's textual syntax with expression and pattern literal forms for XHTML elements \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} gives examles of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2). 
The desugarings of these literals, not shown, are substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.

Third-party library providers are quite justifiably envious of the privileged status of semantically ordinary types like those just discussed that have been equipped with convenient literal forms by a language designer. After all, it is not difficult to find other examples of types for which programmers have invented specialized syntax. For example, 1) clients of a ``collections'' library might like vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur/Web programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}. Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use unstructured string encodings (a.k.a. ``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. \citet{TSLs} sampled string literals from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g. for SQL queries, regexes, file system paths, URLs, IP addresses and others.) Taken together, there appears to be a clear need for mechanisms that allow library providers to introduce new type-specific literal forms into a general-purpose language in a decentralized manner.

Unfortunately, the mechanisms available today are, to put it pithily, \emph{unreasonable}. The next two subsections flesh out this claim and thereby motivate the specific contribution of this paper, introduced in Sec. \ref{sec:contributions}: a decidedly more reasonable mechanism for defining new literal forms within an ML- or Scala-like general-purpose language (i.e. a language that has a rich static type system, support for pattern matching and an ML-style module system or Scala-style object system with support for type members.)

\subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
One well-worn approach is for library providers to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new derived forms (i.e. forms given meaning by expansion, rather than directly.)

\begin{figure}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-6px}
\caption{XHTML literals with support for splicing at two different types are built in to Ur/Web.}
\vspace{-6px}
\label{fig:urweb}
\end{figure}
 % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
% Syntax definition systems , which we will discuss in Sec. \ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a. ``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.

% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}. However, we argue that in fact 
There are hundreds of syntax definition systems of various design, e.g. grammar-based systems like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and somewhat less general systems built in to certain programming systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart} or more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}. 
%library-specific (a.k.a. ``domain-specific'') 
Unfortunately, this dialect-oriented approach is difficult to reconcile with the best practices of ``programming in the large'', i.e. developing large programs ``consisting of many small programs (modules), possibly written by different people''   \cite{DeRemer76} whose interactions are mediated by a reasonable type and binding discipline. The problems that tend to arise are summarized below.%a more systematic treatment will follow in  Sec. \ref{sec:syntax-dialects}.

\paragraph{Problem 1: Incompatibility and Syntactic Conflict}
The first problem with the dialect-oriented approach is that clients  cannot always combine different syntax dialects when they want to use derived forms that they define together. This is a significant problem because large programs use many separately defined libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single preconceived ``problem domain''.

In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems. In other cases, there is no dialect combination operation available that ensures that the combined dialect is an extension of both constituent dialects. And in other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings.

For example, consider two syntax dialects defined under a system like Camlp4: $\mathcal{D}_1$ defines derived forms for sets, and $\mathcal{D}_2$ defines derived forms for dictionaries, both delimited by \verb~{<~ and \verb~>}~.\footnote{In OCaml, simple curly braces are already reserved by the language for record types and values.} Each dialect is deterministic (i.e. there are no possible syntactic conflicts), $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$, but when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$. In particular, \verb~{<>}~ can desugar to the empty set and the empty dictionary.%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets. Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
 %A third syntax dialect might come along that uses the same forms that $\mathcal{D}_2$ defines, but for ordered finite maps.

\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that ``nearly'' guarantees that determinism is conserved when syntax dialects (of a certain restricted class) are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets. To be confident that the marking tokens used are disjoint, providers must base them on the domain name system (or some other coordinating entity.) Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens (as one might do when, for example, deeply nested modules are used for namespacing in ML.)

\paragraph{Problem 2: Abstract Reasoning About Derived Forms}

Even putting aside the problem of syntactic conflict, there are questions about how \emph{reasonable}  sprinkling library-specific derived forms throughout a large software system might be. 
For example, consider the perspective of a programmer attempting to comprehend (i.e. reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of Standard ML's syntax.

@@

If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties. If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:

\begin{enumerate}
\item \textbf{(Responsibility)} Which constituent dialect defined the derived form that appears on Line 3?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e. expressions of variable form), or parsed in some other way peculiar to this form?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break the program, or change its meaning? In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to build in literal forms for a select few standard data structures, e.g. lists. The problem is that this practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.) Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.

This paper formally details a more reasonable alternative: \emph{typed literal macros} (TLMs). TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning that 1) expansions cannot make any assumptions about the application site context; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced segments. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.  This is the first detailed type-theoretic account of a typed, hygienic syntactic macro system for an ML-like language. % This calculus is the first such formalization of a typed macro system.

@@

For example, in Standard ML (SML), the list expression literal \li{[x, y, z]} desugars to a term  semantically equivalent to \li{Cons(x, Cons(y, Cons(z, Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}.\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the expansion of literal forms is \emph{context independent}. We will return to the concept of context independence throughout this work.} List pattern literals are analagous. Another example is Ur/Web, which extends Ur's textual syntax with expression and pattern literal forms for XHTML elements \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} gives examles of XHTML literals containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2). 
The desugarings of these literals, not shown, are substantially more verbose and obscure.

Third-party library providers are quite justifiably envious of the privileged status of semantically ordinary types that have been equipped with convenient literal forms by the language designer. After all, it is not difficult to find other examples of types for which programmers have invented specialized syntax. For example, 1) clients of a ``collections'' library might like not just list literals, but also vector, matrix, finite set and finite map literals following standard mathematical conventions; 2) clients of a ``web programming'' library in SML might like Ur/Web style XHTML literals; 3) Ur/Web programmers might further like CSS literals, which Ur/Web does not provide; 4) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages of interest; and 5) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}. Research by \citet{Bravenboer:2007:PIA:1289971.1289975} suggests that literal forms for structured encodings of SQL queries and other query types can reduce the temptation to use unstructured string encodings (``stringly-typed programming'') and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on the web today \cite{owasp2017}. \citet{TSLs} sampled string literals from open source projects and found that at least 15\% of them were parseable according to some readily apparent type-specific grammar (e.g. for SQL queries, regexes, file system paths, URLs, IP addresses and others.) Taken together, there appears to be a clear need for mechanisms that allow library providers to introduce new type-specific literal forms into a general-purpose language in a decentralized manner.

Unfortunately, the mechanisms available today are, to put it pithily, \emph{unreasonable}. The next two subsections flesh out this claim and thereby motivate the specific contribution of this paper, introduced in Sec. \ref{sec:contributions}: a decidedly more reasonable mechanism for defining new literals within an ML- or Scala-like general-purpose language (i.e. a language that has a rich static type system, support for pattern matching and an ML-style module system or Scala-style object system with support for type members.)

\subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
One approach available to library providers is to use a syntax definition system to construct a library-specific \emph{syntax dialect}: a new syntax definition that extends the syntax definition given in the language definition with new forms, including literal forms. Ur/Web's syntax (Fig. \ref{fig:urweb}) is one example of a syntax dialect \cite{conf/popl/Chlipala15}.

\begin{figure}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-6px}
\caption{XHTML literals with support for splicing at two different types are built primitively into Ur/Web.}
\vspace{-6px}
\label{fig:urweb}
\end{figure}
 % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
% Syntax definition systems , which we will discuss in Sec. \ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a. ``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.

% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}. However, we argue that in fact 
There are hundreds of syntax definition systems of various design, some integrated deeply into languages and build systems, others operating as standalone preprocessors. Examples include grammar-based systems like Camlp4 \cite{ocaml-manual}\todo{better citation?}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and somewhat less general systems built in to certain programming systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart}\todo{better citation?}~ or more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}. 
%library-specific (a.k.a. ``domain-specific'') 
Unfortunately, the dialect-oriented approach is difficult to reconcile with the best practices of ``programming in the large'', i.e. developing large programs ``consisting of many small programs (modules), possibly written by different people'' \cite{DeRemer76}. The reasoning problems that arise are summarized below.%a more systematic treatment will follow in  Sec. \ref{sec:syntax-dialects}.

\paragraph{Problem 1: Modular Reasoning About Syntactic Determinism}
The first problem with the dialect-oriented approach is that clients cannot always deterministically combine different syntax dialects when they want to use the new forms that they define together. This is a significant problem because large programs use many separately developed libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single preconceived ``problem domain''.

In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems. In other cases, there is no dialect combination operation available. And in other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings. For example, consider two syntax dialects defined under a system like \li{camlp4}: $\mathcal{D}_1$ defines literal forms for sets, and $\mathcal{D}_2$ defines literal forms for finite sets, both delimited by \verb~{<~ and \verb~>}~. Each dialect is deterministic (i.e. there are no possible syntactic conflicts), denoted $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$, but when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$. In particular, \verb~{<>}~ can desugar to the empty set and the empty dictionary. Reasoning about determinism is therefore not modular. %(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets. Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
 %A third syntax dialect might come along that uses the same forms that $\mathcal{D}_2$ defines, but for ordered finite maps.

\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that guarantees that determinism is conserved when syntax dialects of a certain restricted class are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets. To be confident that the marking tokens used are disjoint, providers must base them on the domain name system (or some other coordinating entity.) Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens (as one might do when, for example, deeply nested modules are used for namespacing in ML.)

\paragraph{Problem 2: Abstract Reasoning About Literal Forms}

@@

Even putting aside the problem of syntactic conflict, there are questions about just how reasonable sprinkling many possibly unfamiliar library-specific literal forms  throughout a program may be. For example, consider the perspective of a programmer attempting to comprehend (i.e. reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of Standard ML's syntax. If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties. If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:

\begin{enumerate}
\item \textbf{(Responsibility)} Which constituent dialect defined the literal form that appears on Line 3?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e. expressions of variable form), or parsed in some other way peculiar to this literal form?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break the program, or change its meaning? In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}
